---
title: "LMM"
author: "May"
date: "2/22/2021"
output: 
  html_document:
    # self_contained: false
    toc: true
    toc_float: true
    toc_depth: 4
    keep_tex: yes
    theme: united
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)

```

```{r, include=F}
colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color, 
      x)
  } else x
}
```



# Pregnancies Dataset

```{r}
ubarim = read.csv("datasets/ubarim.csv")
```

```{r}
ubarim = ubarim[,-1]
head(ubarim)
```
## Descriptive Statistics

```{r}
ubarim %>% ggplot(aes( x= head_cir  )) +
  geom_histogram( fill = 'darkblue', color = 'white')
```



**NA's** 

```{r}
sum_is_na = function(col) return(sum(is.na(col)))
apply(ubarim,2, sum_is_na)
```

```{r}
dim(ubarim)
```

**How many technitions?**
```{r}
ubarim %>% select(technition) %>% n_distinct()
```

Let's look at the distribution of the number of examinations per technician:
```{r}
ubarim %>% group_by(technition) %>% count() %>%
  ggplot(aes(x=n)) +
  geom_histogram(bins = 25, fill = 'darkblue', color = 'white') +
  labs(x= "No. of examination for a technician",
       title = "Histogram of No. of Examinations per Technician")
```




```{r}
ubarim %>% group_by(technition) %>% count(sort=T)
```

How many mom's?
```{r}
ubarim %>% select(snumber) %>% n_distinct()
```

Distribution of births per mother:
```{r}
ubarim %>% group_by(snumber) %>% summarise(n_births =  n_distinct(meshoar)) %>%
  ggplot(aes(x=factor(n_births))) +
  geom_bar(fill = 'darkblue' , color = 'white') +
  labs( x= 'No. births per mom' , y= 'Birth Count' )
```

```{r}
ubarim %>% group_by(snumber) %>% summarise(n_births =  n_distinct(meshoar)) %>% arrange(-n_births)
```

```{r}
ubarim = ubarim %>% unite("mom_embryo_unique", c("snumber", "meshoar"), remove = FALSE) 
```

How many pregnancies? 
```{r}
ubarim %>% select(mom_embryo_unique) %>% n_distinct()
```

How many pregnancied by technician? 
```{r}
ubarim %>%  group_by(technition) %>% summarise(n_preg = n_distinct(mom_embryo_unique)) %>%
  ggplot(aes(x=n_preg)) +
  geom_histogram(bins = 30, fill = 'darkblue', color = 'white') +
  labs(x= "No. of pregnancies per technician",
       title = "Histogram of No. of pregnancies per Technician")
  
```








## Linear Models

```{r}
ubarim_fit.1 = lm(head_cir~pregweek, data = ubarim)
summary(ubarim_fit.1)
```


```{r}
ubarim %>% ggplot(aes(x=pregweek, y=head_cir)) +
  geom_point(alpha=.03 ) +
  geom_smooth(method = lm, se = FALSE, colour = '#148F77')
```


```{r}
ubarim_fit.2 = lm(head_cir ~ poly(pregweek,3) , data = ubarim)
summary(ubarim_fit.2)
```

```{r}
ubarim %>% ggplot(aes(x=pregweek, y=head_cir)) +
  geom_point(alpha=.05) +
  stat_smooth(method="lm",  formula=y ~ poly(x, 3, raw=TRUE),  colour = '#148F77')
```


## Linear Random Effects
```{r, warning=F, message=F}
library(lme4)
```

**Random intercept for technition:** 
```{r}
lme.1<- lmer( head_cir ~ (1|technition) + pregweek  , data = ubarim )
summary(lme.1)
```


**Random intercept for embryo:** 
```{r}
lme.2<- lmer( head_cir ~ (1|mom_embryo_unique) + pregweek  , data = ubarim )
summary(lme.2)
```

Random intercept for mother: 
```{r}
lme.3<- lmer( head_cir ~ (1|snumber) + pregweek  , data = ubarim )
summary(lme.3)
```

`r colorize("Random Slope???", "red")`  


# Feedback dataset

```{r, message=F, warning=F}
feedback_long_df = read_csv("datasets/feedback_long_df.csv")
head(feedback_long_df)
```

$$
\begin{aligned}
MVC_{ijk} &= \beta_{oj}+\beta_{1j} time_{k} + feedback_{i} +\varepsilon_{ijk} \\
\beta_{0j} &= \gamma_{00}+u_{oj} \\
\beta_{1j} &= \gamma_{10}+u_{1j}
\end{aligned}
$$


```{r}
model.1 =  lmer( performance ~ feedback + time +  (1+time|id) ,
                 data = (feedback_long_df %>%  # make ref group "no feedback"
                   mutate(feedback = ifelse(feedback == "no feedback" , 0 , feedback) )))
summary(model.1)
```

