---
title: "Multinomial Regression"
author: "May Ben-Hamo"
output: 
  html_document:
    self_contained: false
    toc: true
    toc_float: true
    toc_depth: 4
    keep_tex: yes
    theme: united
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=F, message=F}
library(tidyverse)
library(ggplot2)
```


# Iris dataset

The Iris flower data set or Fisher's Iris data set is a multivariate data set introduced by the British statistician, eugenicist, and biologist Ronald Fisher in his 1936 paper The use of multiple measurements in taxonomic problems as an example of linear discriminant analysis.

The data set consists of 50 samples from each of three species of Iris (Iris setosa, Iris virginica and Iris versicolor). Four features were measured from each sample: the length and the width of the sepals and petals, in centimeters. Based on the combination of these four features, Fisher developed a linear discriminant model to distinguish the species from each other.


```{r}
head(iris)

iris_df = iris %>% rename_all(janitor::make_clean_names)
str(iris_df)
```

```{r}
pairs(iris_df[,-5], pch = 21,  
      bg = c("#1b9e77", "#d95f02", "#7570b3")[unclass(iris_df$species)])
```

```{r}
summary(iris_df)
```

```{r}
plot_hist_by_species = function(col, colname) {
  iris_df %>% ggplot(aes(x = species , y = {{col}}, color = species)) +
  geom_boxplot() +  geom_jitter( alpha = .5, position=position_jitter(w=0.1,h=0.1) ) +
  theme(legend.position = "") + labs(title = colname)
}

p1 = plot_hist_by_species(sepal_width, "Sepal Width")
p2 = plot_hist_by_species(sepal_length, "Sepal Length")
p3 = plot_hist_by_species(petal_length, "Petal Length")
p4 = plot_hist_by_species(petal_width, "Petal Width")


gridExtra::grid.arrange(p1, p2, p3, p4)


```

## Fitting Multinomial Regression 

```{r, warning=F, message=F}
library(VGAM)

model_1 = vglm( species ~  sepal_length  +  sepal_width    + petal_length  +  petal_width, family=multinomial, iris_df)
summary(model_1)
```

```{r}
# reference group:virginica
levels(iris_df$species)
```

## Predicting 


```{r}
preds_probs = predict(model_1, type = "response")
preds = apply(preds_probs, MARGIN = 1, which.max)

species_fun = function(level) return(levels(iris$Species)[level])
preds_class = map_chr(preds, species_fun)

# confusion matrix 
table("Predicted" = preds_class, "Actual" = iris_df$species )
```



# National Election Study


**Description**

10 variable subset of the 1996 American National Election Study. Missing values and "don’t know" responses have been listwise deleted. Respondents expressing a voting preference other than Clinton or Dole have been removed  (Rosenstone, Kinder, and Miller (1997)).  

**Format**

A data frame with 944 observations on the following 10 variables:

- `popul` population of respondent’s location in 1000s of people
- `TVnews` days in the past week spent watching news on TV
- `selfLR` Left-Right self-placement of respondent: an ordered factor with levels extremely liberal,
extLib < liberal, Lib < slightly liberal, sliLib < moderate, Mod < slightly conservative,
sliCon < conservative, Con < extremely conservative, extCon
- `ClinLR` Left-Right placement of Bill Clinton (same scale as selfLR): an ordered factor with levels
extLib < Lib < sliLib < Mod < sliCon < Con < extCon
newhamp 67
- `DoleLR`Left-Right placement of Bob Dole (same scale as selfLR): an ordered factor with levels
extLib < Lib < sliLib < Mod < sliCon < Con < extCon
- `PID` Party identification: an ordered factor with levels strong Democrat, strDem < weak Democrat,
weakDem < independent Democrat, indDem < independent independentindind < indepedent
Republican, indRep < waek Republican, weakRep < strong Republican, strRep
- `age` Respondent’s age in years
- `educ` Respondent’s education: an ordered factor with levels 8 years or less, MS < high school
dropout, HSdrop < high school diploma or GED, HS < some College, Coll < Community
or junior College degree, CCdeg < BA degree, BAdeg < postgraduate degree, MAdeg
`income` Respondent’s family income: an ordered factor with levels $3Kminus < $3K-$5K < $5K-$7K
< $7K-$9K < $9K-$10K < $10K-$11K < $11K-$12K < $12K-$13K < $13K-$14K < $14K-$15K <
$15K-$17K < $17K-$20K < $20K-$22K < $22K-$25K < $25K-$30K < $30K-$35K < $35K-$40K
< $40K-$45K < $45K-$50K < $50K-$60K < $60K-$75K < $75K-$90K < $90K-$105K < $105Kplus
vote Expected vote in 1996 presidential election: a factor with levels Clinton and Dole


**Source**: Faraway Book: 



```{r, warning=F, message=F}
library(faraway)
data(nes96)
str(nes96)
```


```{r}
summary(nes96)
```

The response will be party identification of the respondent: Democrat, Independent or Republican. The original data involved more than three categories; They collapse this to three for simplicity.

```{r}
nes96$PID_new = case_when(
  nes96$PID %in% c("strDem", "weakDem") ~ "Democrat",
  nes96$PID %in% c("indDem", "indind", "indRep") ~ "Independent",
  nes96$PID %in% c("weakRep", "strRep") ~ "Republican"
)

table(nes96$PID_new)
```

For simplicity, they consider only the age, education level and income group of the respondents.

```{r}
# unique(nes96$income)


nes96 %>% group_by(income) %>% mutate(sum_inc = n()) %>% ungroup() %>% group_by(income, PID_new) %>%
  summarise(n = n(), pct = n/ sum_inc) %>%
  ggplot( aes( x= income, y= pct, col = PID_new, group = PID_new)) +
  geom_point() + geom_line(alpha = .5, size =1.5) + 
  theme(axis.text.x =  element_text(angle  = 90))
```

In the book they converted income variable to a numberic variable by taking the midpoint of each range, lets do the same ( we don't want to estimate too much parameters):

```{r}
mid_points <- c(1.5,4,6,8,9.5,10.5,11.5,12.5,13.5,14.5,16,18.5,21,23.5,27.5,32.5,37.5,42.5,47.5,55,67.5,82.5,97.5,115)
nes96$numeric_income <- mid_points[unclass(nes96$income)]
summary(nes96$numeric_income)
```



```{r}
nes96 %>% group_by(educ) %>% mutate(sum_inc = n()) %>% ungroup() %>% group_by(educ, PID_new) %>%
  summarise(n = n(), pct = n/ sum_inc) %>%
  ggplot( aes( x= educ, y= pct, col = PID_new, group = PID_new)) +
  geom_point() + geom_line(alpha = .5, size =1.5) 
```

```{r}
nes96 %>% ggplot( aes(  age,  fill = PID_new, color = PID_new)) +
  geom_histogram(aes(y=..density..) , bins = 15, alpha = .8,  color = 'white', position = 'dodge') + 
  geom_density(alpha=.15, fill = 'white')


```



```{r, warning=F, message=F}
library(VGAM)

model_2 = vglm( PID_new ~  age  +  numeric_income     , family=multinomial, nes96)
summary(model_2)
```

```{r}
# ref category - republican
levels(factor(nes96$PID_new))
```

```{r}
predict(model_2,data.frame(age = 25, numeric_income=0),type="response")
predict(model_2,data.frame(age = 25, numeric_income=median(nes96$numeric_income)),type="response")
predict(model_2,data.frame(age = 25, numeric_income=mean(nes96$numeric_income)),type="response")
predict(model_2,data.frame(age = 50, numeric_income=median(nes96$numeric_income)),type="response")
predict(model_2,data.frame(age = 70, numeric_income=median(nes96$numeric_income)),type="response")
```



```{r}
model_3 = vglm( PID_new ~  age  +  numeric_income + educ    , family=multinomial, nes96)
summary(model_3)
```


